Reasoning About a Machine with Local Capabilities

### 2019-02-05 toplas/paper.tex
Place figure on its own page
Make various small improvements to the TOPLAS paper

Compromising software security is often based on attacks that break programming language properties relied upon by software authors, such as control-flow correctness, local-state encapsulation, etc.
Commodity processors offer little support for defending against such attacks: they offer security primitives with only coarse-grained memory protection and limited compartmentalization scalability.
As a result, defenses against attacks on control-flow correctness and local-state encapsulation are either limited to mitigation of only certain common forms of attacks (leading to an attack-mitigation arms race) and/or rely on techniques like machine code rewriting~\citep{wahbe_efficient_1993,abadi_control-flow_2005}, machine code verification~\citep{morrisett_system_1999}, virtual machines with a native stack~\citep{lindholm_java_2014} or randomization~\citep{forrest_building_1997}.
The latter techniques essentially emulate protection techniques on existing hardware, at the cost of performance, system complexity and/or security.

\emph{Capability machines} are a type of processors that
remediate these limitations with a better security model at
the hardware level. They are based on old ideas~\citep{Carter:1994:HSF:195473.195579,Dennis:1966:PSM:365230.365252,shapiro_eros:_1999},
but have recently received renewed interest; in particular, the CHERI
project has proposed new ideas and ways of tackling practical
challenges like backwards compatibility and realistic OS
support~\citep{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740}. Capability
machines tag every word (in the register file and in memory) to
enforce a strict separation between numbers and capabilities (a kind
of pointers that carry authority). Memory capabilities carry
the authority to read and/or write to a range of memory
locations. There is also a form of \emph{object capabilities}, which represent the
authority to invoke a piece of code without exposing the code's
encapsulated private state (e.g., the M-Machine's enter capabilities or
CHERI's sealed code/data pairs).

@@

\dominique{I think we should include a comparison with the conference paper, and explain what the delta is for this journal version.}
In this paper, we make two main contributions: (1) an alternative calling
convention that uses local capabilities to enforce stack frame encapsulation and
well-bracketed control flow, and (2) perhaps more importantly, we adapt and
apply the well-studied techniques of step-indexed Kripke logical relations for
reasoning about code on a representative capability machine with local
capabilities in general and correctness and security of the calling convention
in particular. More specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities and its operational semantics
  (\sectionname~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (\sectionname~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions).
  %It also does not use separate per-component stacks but a single, fairly standard, contiguous stack.
  %domi: don't mention this here. I think some CSF reviewers misunderstood that we were claiming this as a strength of the approach
  It supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel step-indexed Kripke logical relation for reasoning
  about programs on the capability machine. It is an untyped logical relation,
  inspired by previous work on object capabilities \citep{Devriese:2016ObjCap}.
  We prove an analogue of the standard fundamental theorem of logical relations
  --- to the best of our knowledge, our theorem is % by far
  the most general and powerful formulation of the formal guarantees offered by
  a capability machine (a form of capability
  safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
  We are the first to apply these techniques for reasoning about capability
  machines and we believe they will prove useful for many other purposes than
  our calling convention.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem: the use of a
  \emph{single} orthogonal closure (rather than the earlier used biorthogonal
  closure) and a variant of \citet{Dreyer:jfp12}'s public and private future
  worlds~\citep{Dreyer:jfp12} to express the special nature of local
  capabilities. The logical relation and the fundamental theorem expressing
  capability safety are presented in \sectionname~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (\sectionname~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

\begin{toplas}
  This paper is an extension of the published conference paper~\citet{Skorstengaard:esop18}.
  We have made improvements to readability and completeness throughout the paper, but we highlight the following changes:
  \begin{itemize}
  \item An introduction to the \sectionname~\nameref{sec:logical-relation} that provides informal intuition about how the logical relation machinery comes into play on a capability machine.
  \item Proof sketches for the \nameref{thm:ftlr} and Lemma~\ref{lem:correctness-g1}, the correctness lemma for the awkward example.
  \item Figures that illustrate central parts of the calling convention.
  \item A section about malloc.
    Specifically, we provide the specification for the malloc used in the examples of the paper.
  \item A section about macro instructions with descriptions of all the macros and the implementation of \texttt{scall}.
  \item A section on reasoning about programs that run on a capability machine.
    This section explains how one reasons about common scenarios that arise in programs on a capability machine, and in particular how the logical relation is used.
    It also introduces a number of lemmas that prove recurring bits once and for all.
  \item Details previously found only in the technical appendix~\citep{technical_appendix} to the paper.
    Further, many of the still omitted details have been added to the paper's appendix.
  \end{itemize}
\end{toplas}

We have written a technical appendix~\citep{technical_appendix} which contains
additional details and proofs left out from this paper.

@@

We assume a finite set of register names $\RegName$.
We define register files $\reg$ and memories $\ms$ as functions mapping register names resp.\ addresses to words.
The state of the entire machine is represented as a configuration that is either a running state $\Phi \in \ExecConfs$ containing a memory and a register file or a failed or halted state, where the latter is paired with the final state of memory.

@@

The $\jmp{}$ instruction updates the program counter to a requested location,
but it is complicated by the presence of \emph{enter capabilities} 
after the M-Machine's~\citep{Carter:1994:HSF:195473.195579}. Enter capabilities
cannot be used to read, write or execute and their address and range cannot be
modified. They can only be used to jump to, but when that happens, their
permission changes to $\exec$. They can be used to represent a kind of closures:
an opaque package containing a piece of code together with local encapsulated
state. Such a package can be built as an enter capability $c =
((\entry,\gl),\start,\addrend,\addr)$ where the range $[\start,\addr-1]$
contains local state (data or capabilities) and $[\addr,\addrend]$ contains
instructions. The package is opaque to an adversary holding $c$ but when $c$ is
jumped to, the instructions can start executing and have access to the local
data through the updated version of $c$, now in the $\pcreg$-register. 
\lau{As we talked about what is described here, is not how we make closures in
the TR, so we should consider whether we should refrain from using that word
here. }

@@

Finally, the $\store{}{}$ instruction updates the memory to the requested value
if the capability has write authority for the requested location. However, the
instruction is complicated by the presence of \emph{local capabilities}, modeled
after the ones in the CHERI processor~\citep{Watson2015Cheri}. Basically, local
capabilities are special in that they can only be kept in registers, i.e.\ they
cannot be stored to memory. This means that local capabilities can be
\emph{temporarily} given to an adversary, for the duration of an invocation: if
we make sure to clear the capability from the register file after control is
passed back to us, they will not have been able to store the capability.
However, there is one exception to the rule above: local capabilities can be
stored to memory for which we have a capability with write-local authority
(i.e.\ permission $\rwl$ or $\rwlx$). This is intended to accommodate a stack,
where register contents can be stored, including local capabilities. As long as
all capabilities with write-local authority are themselves local and the stack
is cleared after control is passed back by the adversary, we will see that this
does not break the intended behavior of local capabilities.

We point out that our local capabilities capture only a part of the semantics of
local capabilities in CHERI. Specifically, in addition to the above, CHERI's
default implementation of the CCall exception handler forbids local capabilities
from being passed across module boundaries. Such a restriction fundamentally
breaks our calling convention, since we pass around local return pointers and
stack capabilities. However, CHERI's CCall is not implemented in hardware, but
in software, precisely to allow experimenting with alternative models like ours.
\dominique{Update to CHERI's new ccall story.}

@@

The basic idea of our approach is simple: we stick to a single, rather standard,
C stack and register-passed stack and return pointers, much like a standard C
calling convention. However, to prevent various ways of misusing this basic
scheme, we put local capabilities to work and take a number of
not-always-obvious safety measures. The safety measures are presented in terms
of what \emph{we} need to do to protect ourselves against an \emph{adversary},
but this is only for presentation purposes as our code assumes no special status on
the machine. In fact, an adversary can apply the same safety measures to protect
themselves against us. In the following paragraphs, we will explain the issues to be
considered in all the relevant situations: when (1) starting our program, (2)
returning to the adversary, (3) invoking the adversary, (4) returning from the
adversary, (5) invoking an adversary callback and (6) having a callback invoked
by the adversary.

@@

However, if we are returning from a function that has itself invoked adversary
code \itoplas{(\figurename~\ref{fig:illu-call-adv})}, then clearing the used part of the stack is not enough. The \emph{unused}
part of the stack may also contain data and capabilities, left there by the
adversary, including local capabilities since the stack is write-local. As we
will see later, we rely on the fact that the adversary cannot keep hold of local
capabilities when they pass control to the trusted code and receive control
back. In this case, the adversary could use the unused part of the stack to
store local pointers and load them from there after they get control back
\itoplas{(\figurename~\ref{fig:illu-adv-data})}. To
prevent this, we need to clear (i.e.\ overwrite with zeros) the entire part of
the stack that the adversary has had access to, not just the parts that we have
used ourselves \itoplas{(\figurename~\ref{fig:illu-clear-all})}. Since we may be talking about a large part of memory, this
requirement is the most problematic aspect of our calling convention for
performance, but see \sectionname~\ref{sec:discussion} for how this might be
mitigated.

\begin{figure}[tb]
  \centering
  \begin{subfigure}[t]{0.3\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

      % Stack pointer
      \begin{scope}
        \clip (4.6,2) rectangle (7.6,11.5);
        \capbrace{(4.6,2)}{(4.6,12)}{sp}
      \end{scope}
    \end{tikzpicture}
    \caption{The stack just after we returned to the adversary and did take
      care to clear our local stack frame.}
  \label{fig:illu-clear-priv}
  \end{subfigure}
  \begin{subfigure}[t]{0.33\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

      % Stack pointer
      \begin{scope}
        \clip (4.6,5) rectangle (7.6,11.5);
        \capbrace{(4.6,5)}{(4.6,13)}{sp}
      \end{scope}
    \end{tikzpicture}
  \caption{The stack after an adversary has called us and we have called the adversary.}
  \label{fig:illu-call-adv}
\end{subfigure}
\begin{subfigure}[t]{0.043\linewidth}
  % padding
  \phantom{a}
\end{subfigure}
  \begin{subfigure}[t]{0.3\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

In one of the registers, we also provide a return pointer which must be a local
capability. If it were global, the adversary would be able to store away the
return pointer in a global data structure (i.e.\ there exists a global capability for it), and jump to it later, in
circumstances where this should not be possible. For example, they could store
the return pointer, legally jump to it a first time, wait to be invoked again,
and then jump to the old return pointer a second time, instead of the new return
pointer received for the second invocation. Similarly, they could store the
return pointer, invoke a function in our code, wait for us to invoke them again,
and then jump to the old return pointer rather than the new one, received for the
second invocation.
% \lau{If we need space, then maybe one of the examples is sufficient.}
% \lau{Both of these are good to have their as the two last examples of \sectionname~\ref{sec:examples} use them.}
%
% \lau{In the examples, we refer to the local return pointer as
%   ``protected return pointer''. We should either introduce that here
%   or use a different formulation in the examples. }
% \lau{It seemed did not seem to add anything to adopt the terminlogy - we
%   often highlight that the return pointer is local which in most cases
%   is so important that we do not want to leave it out.}
By making the return pointer local, we prevent such attacks.
The adversary can only store local capabilities with a write-local capability, and the only piece of memory governed by a write-local capability is the stack.
Since the stack pointer itself is also local, it can also only be stored on the stack.
There is no way for the adversary to recover either of these local capabilities because we clear the part of the stack that they had access to before we pass control back to them.

@@

    % attacker ..
      \actadv{(0,2)}{(4.5,5)}{Adv. stack frame}
      \draw[fill=white] (0,10) rectangle (4.5,10.5) node[pos=.5,color=red] {\footnotesize sp};
    % Stack pointer
      \begin{scope}
        \clip (4.6,2) rectangle (7.6,11.5);
        \capbracebot{(4.6,2)}{(4.6,12)}{sp}
      \end{scope}
    \end{tikzpicture}
    \caption{The stack after the adversary has its stored the stack pointer in
      far up on the stack.}
  \label{fig:illu-adv-store-sp}
\end{subfigure}
\begin{subfigure}{0.043\linewidth}
  % padding
  \phantom{a}
\end{subfigure}
  \begin{subfigure}[t]{0.3\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

    \end{tikzpicture}
  \caption{The stack after the adversary have called some trusted code.}
  \label{fig:illu-adv-call}
  \end{subfigure}
\begin{subfigure}{0.043\linewidth}
  % padding
  \phantom{a}
\end{subfigure}
  \begin{subfigure}[t]{0.3\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

    % Stack pointer
      \begin{scope}
        \clip (4.6,2) rectangle (7.6,11.5);
        \capbracebotred{(4.6,2)}{(4.6,12)}{sp}
      \end{scope}
      \begin{scope}
        \clip (5.1,6.6) rectangle (9,11.5);
        \capbrace{(5.1,6.5)}{(5.1,12)}{sp'}
      \end{scope}
    \end{tikzpicture}
  \caption{The stack after the trusted code has called the adversary. The
    adversary's old stack pointer is still stored on the stack.}
  \label{fig:illu-adv-called}
  \end{subfigure}
  \caption{Illustration of the situations related to stack clearing when invoking an adversary.
    The greyed out areas are parts of the stack that are not accessible at the time.
  }
  \label{fig:invoke-adv}
\end{figure}

\textbf{Returning from the adversary} So return pointers must be passed as local
capabilities. But what should their permissions be, what memory should they
point to and what should that memory (the activation record) contain? Let us
answer the last question first by considering what should happen when the
adversary jumps to a return pointer. In that case, the program counter should be
restored to the instruction after the jump to the adversary, so the activation
record should store this old program counter. Additionally, the stack pointer
should also be restored to its original value. Since the adversary has a more
restricted authority over the stack than the code making the call, we cannot
hope to reconstruct the original stack pointer from the stack pointer owned by the
adversary. Instead, it should be stored as part of the activation record.

@@

Clearly, neither the old program counter or the old stack pointer should be accessible by the adversary. In
other words, the return pointer provided to the adversary must be a capability
that they can jump to but not read from, i.e.\ an enter capability. To make this
work, we construct the activation record as depicted in
\figurename~\ref{fig:activ-rec-struct}. The $\entry$ return pointer has authority
over the entire activation record (containing the previous return and stack
pointer), and its current address points to a number of restore instructions in
the record, so that upon invocation, these instructions are executed and can
load the old stack pointer and program counter back into the register file. As
the return pointer is an enter pointer, the adversary cannot get hold of the
activation record's contents, but after invocation, its permission is updated to
$\exec$, so the contents become available to the restore instructions.

The final question that remains is: where should we store this activation
record? The attentive reader may already see that there is only one possibility:
since the activation record contains the old stack pointer, which is local, the
activation record can only be constructed in a part of memory where we have
write-local access, i.e.\ on the stack. Note that this means we will be placing
and executing instructions on the stack, i.e.\ it will not just contain code
pointers and data. This means that our calling convention should be combined
with protection against stack smashing attacks (i.e.\ buffer overflows on the
stack overwriting activation records' contents). Luckily, the capability
machine's fine-grained memory protection should make it reasonably easy for a
compiler to implement such protection, by making sure that only appropriately
bounded versions of the stack pointer are made available to source language
code.

@@

Our answer is that the adversary should provide it to us, just as we provide
them with a stack pointer when we invoke their code. However, it is important
that we do not just accept any capability as a stack pointer but check that it
is safe to use. Specifically, we check that it is indeed an $\rwlx$ capability. Without
this check, an adversary could potentially get control over our local stack
frame during a subsequent callback by passing us a local $\rwx$ capability to a global data
structure instead of a proper stack pointer
and a global callback for our callback to invoke. If our local state contains no
local capabilities, then, otherwise following our calling convention, the callback would
not fail and the adversary could use a stored capability for the global data
structure to access our local state. To prevent this from happening, we need to
make sure the stack capability carries $\rwlx$ authority, since the system wide assumption then
tells us that the adversary cannot have global capabilities to our local stack.

@@

In this section, we construct such a formalization. We make use of some
well-studied and powerful (but non-trivial) machinery from the literature.
Specifically, we employ a unary step-indexed Kripke logical relation with
recursive worlds, and some additional special characteristics of our own.
Step-indexing, Kripke logical relations and recursive worlds are techniques that
may be familiar from lambda calculus settings, but it may not be clear to the
reader how they apply in this more low-level assembly language. Therefore, in the next section, we do
not immediately dive into the details, but we first try to provide some informal
intuition about how all of this machinery comes into play in our setting.

>>>>

Compromising software security is often based on attacks that break programming language properties relied upon by software authors such as control-flow correctness, local-state encapsulation, etc.
Commodity processors offer little support for defending against such attacks: they offer security primitives with only coarse-grained memory protection and limited compartmentalization scalability.
As a result, defenses against attacks on control-flow correctness and local-state encapsulation are either limited to mitigation of only certain common forms of attacks (leading to an attack-mitigation arms race) and/or rely on techniques like machine code rewriting~\citep{wahbe_efficient_1993,abadi_control-flow_2005}, machine code verification~\citep{morrisett_system_1999}, virtual machines with a native stack~\citep{lindholm_java_2014} or randomization~\citep{forrest_building_1997}.
The latter techniques essentially emulate protection techniques on existing hardware at the cost of performance, system complexity, and/or security.

\emph{Capability machines} are a type of processors that
remediate these limitations with a better security model at
the hardware level. They are based on old
ideas~\citep{Carter:1994:HSF:195473.195579,Dennis:1966:PSM:365230.365252,shapiro_eros:_1999} that
 have recently received renewed interest. in particular, the CHERI
project has proposed new ideas and ways of tackling practical
challenges like backwards compatibility and realistic OS
support~\citep{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740}. Capability
machines tag every word (in the register file and in memory) to
enforce a strict separation between numbers and capabilities (a kind
of pointers that carry authority). Memory capabilities carry
the authority to read and/or write to a range of memory
locations. There is also a form of \emph{object capabilities}, which represent the
authority to invoke a piece of code without exposing the code's
encapsulated private state (e.g., the M-Machine's enter capabilities or
CHERI's sealed code/data pairs).

@@

\dominique{I think we should include a comparison with the conference paper, and explain what the delta is for this journal version.}
In this paper, we make two main contributions: (1) an alternative calling
convention that uses local capabilities to enforce stack frame encapsulation and
well-bracketed control flow, and (2) perhaps more importantly, we adapt and
apply the well-studied techniques of step-indexed Kripke logical relations for
reasoning about code on a representative capability machine with local
capabilities in general and correctness and security of the calling convention
in particular. More specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities and its operational semantics
  (\sectionname~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (\sectionname~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions).
  %It also does not use separate per-component stacks but a single, fairly standard, contiguous stack.
  %domi: don't mention this here. I think some CSF reviewers misunderstood that we were claiming this as a strength of the approach
  It supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel step-indexed Kripke logical relation for reasoning
  about programs on the capability machine. It is an untyped logical relation,
  inspired by previous work on object capabilities \citep{Devriese:2016ObjCap}.
  We prove an analogue of the standard fundamental theorem of logical relations
  --- to the best of our knowledge, our theorem is % by far
  the most general and powerful formulation of the formal guarantees offered by
  a capability machine (a form of capability
  safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
  We are the first to apply these techniques for reasoning about capability
  machines and we believe they will prove useful for many other purposes than
  our calling convention.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem: the use of a
  \emph{single} orthogonal closure (rather than the earlier used biorthogonal
  closure) and a variant of \citet{Dreyer:jfp12}'s public and private future
  worlds~\citep{Dreyer:jfp12} to express the special nature of local
  capabilities. The logical relation and the fundamental theorem expressing
  capability safety are presented in \sectionname~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (\sectionname~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}
\begin{toplas}
  This paper is an extension of the published conference paper~\citet{Skorstengaard:esop18}.
  We have made improvements to readability and completeness throughout the paper, but we highlight the following changes:
  \begin{itemize}
  \item An introduction to the \sectionname~\nameref{sec:logical-relation} that provides informal intuition about how the logical relation machinery comes into play on a capability machine.
  \item Proof sketches for the \nameref{thm:ftlr} and Lemma~\ref{lem:correctness-g1}, the correctness lemma for the awkward example.
  \item Figures that illustrate central parts of the calling convention.
  \item A section about malloc.
    Specifically, we provide the specification for the malloc used in the examples of the paper.
  \item A section about macro instructions with descriptions of all the macros and the implementation of \texttt{scall}.
  \item A section on reasoning about programs that run on a capability machine.
    This section explains how one reasons about common scenarios that arise in programs on a capability machine, and in particular how the logical relation is used.
    It also introduces a number of lemmas that prove recurring bits once and for all.
  \item Details previously found only in the technical appendix~\citep{technical_appendix} to the paper.
    Further, many of the still omitted details have been added to the paper's appendix.
  \end{itemize}
\end{toplas}
We have written a technical appendix~\citep{technical_appendix} which contains
additional details and proofs left out from this paper.

@@

We assume a finite set of register names $\RegName$.
We define register files $\reg$ and memories $\ms$ as functions mapping register names resp.\ addresses to words.
The state of the entire machine is represented as a configuration that is either a running state $\Phi \in \ExecConfs$ containing a memory and a register file or a failed or halted state where the latter is paired with the final state of memory.

@@

The $\jmp{}$ instruction updates the program counter to a requested location,
but it is complicated by the presence of \emph{enter capabilities} 
after the M-Machine's~\citep{Carter:1994:HSF:195473.195579}. Enter capabilities
cannot be used to read, write or execute and their address and range cannot be
modified. They can only be used to jump to. When that happens, their
permission changes to $\exec$. They can be used to represent a kind of closures:
an opaque package containing a piece of code together with local encapsulated
state. Such a package can be built as an enter capability $c =
((\entry,\gl),\start,\addrend,\addr)$ where the range $[\start,\addr-1]$
contains local state (data or capabilities) and $[\addr,\addrend]$ contains
instructions. The package is opaque to an adversary holding $c$. When $c$ is
jumped to however, the instructions can start executing and have access to the local
data through the updated version of $c$, now in the $\pcreg$-register. 
\lau{As we talked about what is described here, is not how we make closures in
the TR, so we should consider whether we should refrain from using that word
here. }

@@

Finally, the $\store{}{}$ instruction updates the memory to the requested value
if the capability has write authority for the requested location. However, the
instruction is complicated by the presence of \emph{local capabilities} modeled
after the ones in the CHERI processor~\citep{Watson2015Cheri}. Basically, local
capabilities are special in that they can only be kept in registers, i.e.\ they
cannot be stored to memory. This means that local capabilities can be
\emph{temporarily} given to an adversary, for the duration of an invocation: if
we make sure to clear the capability from the register file after control is
passed back to us, they will not have been able to store the capability.
However, there is one exception to the rule above: local capabilities can be
stored to memory for which we have a capability with write-local authority
(i.e.\ permission $\rwl$ or $\rwlx$). This is intended to accommodate a stack,
where register contents can be stored, including local capabilities. As long as
all capabilities with write-local authority are themselves local and the stack
is cleared after control is passed back by the adversary, we will see that this
does not break the intended behavior of local capabilities.

We point out that our local capabilities capture only a part of the semantics of
local capabilities in CHERI. Specifically in addition to the above, CHERI's
default implementation of the CCall exception handler forbids local capabilities
from being passed across module boundaries. Such a restriction fundamentally
breaks our calling convention since we pass around local return pointers and
stack capabilities. However, CHERI's CCall is not implemented in hardware but
in software precisely to allow experimenting with alternative models like ours.
\dominique{Update to CHERI's new ccall story.}

@@

The basic idea of our approach is simple: we stick to a single, rather standard,
C stack and register-passed stack and return pointers much like a standard C
calling convention. However, to prevent various ways of misusing this basic
scheme, we put local capabilities to work and take a number of
not-always-obvious safety measures. The safety measures are presented in terms
of what \emph{we} need to do to protect ourselves against an \emph{adversary},
but this is only for presentation purposes as our code assumes no special status on
the machine. In fact, an adversary can apply the same safety measures to protect
themselves against us. In the following paragraphs, we will explain the issues to be
considered in all the relevant situations: when (1) starting our program, (2)
returning to the adversary, (3) invoking the adversary, (4) returning from the
adversary, (5) invoking an adversary callback, and (6) having a callback invoked
by the adversary.

@@

However, if we are returning from a function that has itself invoked adversary
code \itoplas{(\figurename~\ref{fig:illu-call-adv})}, then clearing the used part of the stack is not enough. The \emph{unused}
part of the stack may also contain data and capabilities, left there by the
adversary, including local capabilities since the stack is write-local. As we
will see later, we rely on the fact that the adversary cannot keep hold of local
capabilities when they pass control to the trusted code and receive control
back. In this case, the adversary could use the unused part of the stack to
store local pointers and load them from there after they get control back
\itoplas{(\figurename~\ref{fig:illu-adv-data})}. To
prevent this, we need to clear (i.e.\ overwrite with zeros) the entire part of
the stack that the adversary has had access to; not just the parts that we have
used ourselves \itoplas{(\figurename~\ref{fig:illu-clear-all})}. Since we may be talking about a large part of memory, this
requirement is the most problematic aspect of our calling convention for
performance (see the discussion in \sectionname~\ref{sec:discussion} for how this might be
mitigated).

\begin{figure}[ptb]
  \centering
  \begin{subfigure}[t]{0.3\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

      % Stack pointer
      \begin{scope}
        \clip (4.6,2) rectangle (7.6,11.5);
        \capbrace{(4.6,2)}{(4.6,12)}{sp}
      \end{scope}
    \end{tikzpicture}
    \caption{The stack just after we returned to the adversary, and we did take
      care to clear our local stack frame.}
  \label{fig:illu-clear-priv}
  \end{subfigure}
  \begin{subfigure}[t]{0.33\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

      % Stack pointer
      \begin{scope}
        \clip (4.6,5) rectangle (7.6,11.5);
        \capbrace{(4.6,5)}{(4.6,13)}{sp}
      \end{scope}
    \end{tikzpicture}
  \caption{The stack after an adversary has called us, and we have called the adversary.}
  \label{fig:illu-call-adv}
\end{subfigure}
\begin{subfigure}[t]{0.043\linewidth}
  % padding
  \phantom{a}
\end{subfigure}
  \begin{subfigure}[t]{0.3\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

In one of the registers, we also provide a return pointer which must be a local
capability. If it were global, the adversary would be able to store away the
return pointer in a global data structure (i.e.\ there exists a global capability for it) and jump to it later in
circumstances where this should not be possible. For example, they could store
the return pointer, legally jump to it a first time, wait to be invoked again,
and then jump to the old return pointer a second time instead of the new return
pointer received for the second invocation. Similarly, they could store the
return pointer, invoke a function in our code, wait for us to invoke them again,
and then jump to the old return pointer rather than the new one received for the
second invocation.
% \lau{If we need space, then maybe one of the examples is sufficient.}
% \lau{Both of these are good to have their as the two last examples of \sectionname~\ref{sec:examples} use them.}
%
% \lau{In the examples, we refer to the local return pointer as
%   ``protected return pointer''. We should either introduce that here
%   or use a different formulation in the examples. }
% \lau{It seemed did not seem to add anything to adopt the terminlogy - we
%   often highlight that the return pointer is local which in most cases
%   is so important that we do not want to leave it out.}
By making the return pointer local, we prevent such attacks.
The adversary can only store local capabilities with a write-local capability, and the only piece of memory governed by a write-local capability is the stack.
Since the stack pointer itself is also local, it can also only be stored on the stack.
There is no way for the adversary to recover either of these local capabilities because we clear the part of the stack that they had access to before we pass control back to them.

@@

    % attacker ..
      \actadv{(0,2)}{(4.5,5)}{Adv. stack frame}
      \draw[fill=white] (0,10) rectangle (4.5,10.5) node[pos=.5,color=red] {\footnotesize sp};
    % Stack pointer
      \begin{scope}
        \clip (4.6,2) rectangle (7.6,11.5);
        \capbracebot{(4.6,2)}{(4.6,12)}{sp}
      \end{scope}
    \end{tikzpicture}
    \caption{The stack after the adversary has stored its stack pointer
      far up on the stack.}
  \label{fig:illu-adv-store-sp}
\end{subfigure}
\begin{subfigure}{0.043\linewidth}
  % padding
  \phantom{a}
\end{subfigure}
  \begin{subfigure}[t]{0.3\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

    \end{tikzpicture}
  \caption{The stack after the adversary has called some trusted code.}
  \label{fig:illu-adv-call}
  \end{subfigure}
\begin{subfigure}{0.043\linewidth}
  % padding
  \phantom{a}
\end{subfigure}
  \begin{subfigure}[t]{0.3\linewidth}
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

    % Stack pointer
      \begin{scope}
        \clip (4.6,2) rectangle (7.6,11.5);
        \capbracebotred{(4.6,2)}{(4.6,12)}{sp}
      \end{scope}
      \begin{scope}
        \clip (5.1,6.6) rectangle (9,11.5);
        \capbrace{(5.1,6.5)}{(5.1,12)}{sp'}
      \end{scope}
    \end{tikzpicture}
  \caption{The stack after the trusted code has called the adversary. The
    adversary's old stack pointer is still available on the stack.}
  \label{fig:illu-adv-called}
  \end{subfigure}
  \caption{Illustration of the situations related to stack clearing when invoking an adversary.
    The greyed out areas are parts of the stack that are not accessible at the time.
  }
  \label{fig:invoke-adv}
\end{figure}

\textbf{Returning from the adversary} Return pointers must be passed as local
capabilities. But what should their permissions be, what memory should they
point to and what should that memory (the activation record) contain? Let us
answer the last question first by considering what should happen when the
adversary jumps to a return pointer. In that case, the program counter should be
restored to the instruction after the jump to the adversary, so the activation
record should store this old program counter. Additionally, the stack pointer
should also be restored to its original value. Since the adversary has a more
restricted authority over the stack than the code making the call, we cannot
hope to reconstruct the original stack pointer from the stack pointer owned by the
adversary. Instead, it should be stored as part of the activation record.

@@

Clearly, neither the old program counter nor the old stack pointer should be accessible by the adversary. In
other words, the return pointer provided to the adversary must be a capability
that they can jump to but not read from, i.e.\ an enter capability. To make this
work, we construct the activation record as depicted in
\figurename~\ref{fig:activ-rec-struct}. The $\entry$ return pointer has authority
over the entire activation record (containing the previous return and stack
pointer) and its current address points to a number of restore instructions in
the record. Upon invocation, the instructions in the activation record are executed and can
load the old stack pointer and program counter back into the register file. As
the return pointer is an enter pointer, the adversary cannot get hold of the
activation record's contents. However after invocation, its permission is updated to
$\exec$, so the contents become available to the restore instructions.

The final question that remains is: where should we store this activation
record? The attentive reader may already see that there is only one possibility:
since the activation record contains the old stack pointer, which is local, the
activation record can only be constructed in a part of memory where we have
write-local access, i.e.\ on the stack. Note that this means we will be placing
and executing instructions on the stack, i.e.\ it will not just contain code
pointers and data. This means that our calling convention should be combined
with protection against stack smashing attacks (i.e.\ buffer overflows on the
stack overwriting activation records' contents). Luckily, the capability
machine's fine-grained memory protection should make it reasonably easy for a
compiler to implement such protection by making sure that only appropriately
bounded versions of the stack pointer are made available to source language
code.

@@

Our answer is that the adversary should provide it to us; just as we provide
them with a stack pointer when we invoke their code. However, it is important
that we do not just accept any capability as a stack pointer but check that it
is safe to use. Specifically, we check that it is indeed an $\rwlx$ capability. Without
this check, an adversary could potentially get control over our local stack
frame during a subsequent callback by passing us a local $\rwx$ capability to a global data
structure instead of a proper stack pointer
and a global callback for our callback to invoke. If our local state contains no
local capabilities, then, otherwise following our calling convention, the callback would
not fail and the adversary could use a stored capability for the global data
structure to access our local state. To prevent this from happening, we need to
make sure the stack capability carries $\rwlx$ authority since the system wide assumption then
tells us that the adversary cannot have global capabilities to our local stack.

@@

In this section, we construct such a formalization. We make use of some
well-studied and powerful (but non-trivial) machinery from the literature.
Specifically, we employ a unary step-indexed Kripke logical relation with
recursive worlds and some additional special characteristics of our own.
Step-indexing, Kripke logical relations, and recursive worlds are techniques that
may be familiar from lambda calculus settings, but it may not be clear to the
reader how they apply in this more low-level assembly language. Therefore, in the next section, we do
not immediately dive into the details, but first we try to provide some informal
intuition about how all of this machinery comes into play in our setting.
